Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-07-04 icfp24paper/paper.tex
cosmetics
camera ready paper

%% \documentclass[acmsmall,screen,review,anonymous]{acmart}
%% \documentclass[nonacm,acmsmall]{acmart}
\documentclass[acmsmall,screen]{acmart}
%% \raggedbottom

@@

%% \setcopyright{rightsretained}
%% \acmPrice{}
%% \acmDOI{10.1145/3547641}
%% \acmYear{2022}
%% \copyrightyear{2022}
%% \acmSubmissionID{icfp22main-p52-p}
%% \acmJournal{PACMPL}
%% \acmVolume{6}
%% \acmNumber{ICFP}
%% \acmArticle{110}
%% \acmMonth{8}

%% These commands are for a JOURNAL article.

%% \acmJournal{JACM}
%% \acmVolume{37}
%% \acmNumber{4}
%% \acmArticle{111}
%% \acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

@@

\title{Closure-Free Functional Programming in a Two-Level Type Theory}

%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

\author{András Kovács}
\orcid{0000-0002-6375-9781}
\email{andrask@chalmers.se}
\affiliation{%
  \institution{University of Gothenburg}
  \country{Sweden}
  \city{Gothenburg}
}

%% \author{András Kovács}
%% \email{andrask@chalmers.se}
%% \orcid{0000-0002-6375-9781}
%% \affiliation{%
%%   \institution{University of Gothenburg}
%%   \country{Sweden}
%%   \city{Gothenburg}
%% }

@@

\section*{Acknowledgements}
I thank the anonymous reviewers, Sebastian Graf and Jacques Carette, for their
comments and suggestions on draft versions of this paper. This work was supported
by grant 2019.0116 of the Knut and Allice Wallenberg Foundation.

@@

\newpage

@@

\subsection{W-types}
So far we have stayed close to \cite{staged2ltt}, with only cosmetic
changes. Supporting W-types, as a way to get inductive families \cite{whynotw},
is the biggest extension here. We extend Section 5 of \cite{staged2ltt}. In the
syntax of CFTT, $\vW$-types are as follows. The formal specification of
syntactic $\vW{-}\msf{elim}$ is rather gnarly, because of the usage of explicit
substitutions, but we write it out here anyways.

\begin{notation} We use $\p^N$ for $N$-fold
weakening and we also use numerals as De Bruijn indices, to abbreviate
$\q[\p^N]$. We write $\msf{app}$ and $\msf{lam}$ for $\Pi$ application and
abstraction. We write $\blank{\Rightarrow}\blank$ for non-dependent function
types. Finally, for the sake of brevity we omit the component maps of the isomorphism
$\re_\ext : \re\,(\Gamma \ext A) \simeq ((\gamma : \re\,\Gamma) \times \re\,A\,\gamma)$.
\end{notation}
\begin{alignat*}{3}
  &\vW   &&: (A : \MTy_i\,\Gamma) \to \MTy_j\,(\Gamma\,\ext\,A) \to \MTy_{\msf{max}(i,j)}\,\Gamma\\
  &\vsup &&: (a : \MTm\,\Gamma\,A) \to \MTm\,\Gamma\,(B[\vid,\,a] \Rightarrow \vW\,A\,B) \to \MTm\,\Gamma\,(\vW\,A\,B)\\
  &\vW{-}\msf{elim}  &&: (P : \MTy\,(\Gamma\,\ext\,\vW\,A\,B))\\
  &                  && \hspace{0.6em}\to \MTm\,(\Gamma\,\ext\,A\,\ext\,B \Rightarrow (\vW\,A\,B)[p]\,\ext\,\Pi\,(B[p])\,(P[\p^3,\,\msf{app}\,1\,0]))\,(P[\p^3,\,\vsup\,2\,1))\\
  &                  && \hspace{0.6em}\to (w : \MTm\,\Gamma\,(\vW\,A\,B)) \to \MTm\,\Gamma\,(P[\id,\,w])\\
  & \vW{\beta}       &&:
    \vW{-}\msf{elim}\,P\,s\,(\vsup\,a\,f) = s[\id,\,a,\,f,\,\msf{lam}\,(\vW{-}\msf{elim}\,(P[\p^2,\,\q])\,(s[\p])\,(\msf{app}\,f\,0)]
\end{alignat*}
We shall omit the $i$ and $j$ universe levels in the following. In $\whset$,
i.e.\ internally to presheaves over $\mbbo$, we also have the standard W-type.
\begin{alignat*}{3}
  & \wh{\vW}   &&: (A : \whset) \to (A \to \whset) \to \whset\\
  & \wh{\vsup} &&: (a : A) \to (B\,a \to \wh{\vW}\,A\,B) \to \wh{\vW}\,A\,B\\
  & \wh{\vW{-}\msf{elim}} &&: (P : \wh{\vW}\,A\,B \to \whset)\\
  & && \hspace{0.6em}\to ((a : A)(f : B\,a \to \wh{\vW}\,A\,B) \to ((b : B\,a) \to P\,(f\,b)) \to P\,(\wh{\vsup}\,a\,f))\\
  & && \hspace{0.6em}\to (w : \wh{\vW}\,A\,B) \to P\,w\\
  & \wh{\vW}{\beta} &&: \wh{\vW{-}\msf{elim}}\,P\,s\,(\wh{\vsup}\,a\,f) = s\,a\,f\,(\lambda\,b.\,\wh{\vW{-}\msf{elim}}\,P\,s\,(f\,b))
\end{alignat*}
In the presheaf model we interpret $\vW$ using $\whW$.
\begin{alignat*}{3}
  & \ev\,(\vW\,A\,B)\,\gamma   &&:= \wh{\vW}\,(\ev\,A\,\gamma)\,(\lambda\,\alpha.\,\ev\,B\,(\gamma,\,\alpha))\\
  & \ev\,(\vsup\,a\,f)\,\gamma &&:= \wh{\vsup}\,(\ev\,a\,\gamma)\,(\lambda\,\beta.\,\ev\,f\,(\gamma, \beta))\\
  & \ev\,(\vW{-}\msf{elim}\,P\,s\,w)\,\gamma &&:= \wh{\vW{-}\msf{elim}}\,(\lambda\,w.\,\ev\,P\,(\gamma,\,w))\,(\lambda\,a\,f\,g.\,\ev\,s\,(\gamma,\,a,\,f,\,g))\,(\ev\,w\,\gamma)
\end{alignat*}

%% W-elim̂ (λ w. E P (γ, w)) (λ a f g. E s (γ, a, f, g)) (E w γ)
Now, assuming $A : \MTy\,\Gamma$ and $B : \MTy\,(\Gamma\,\ext\,A)$, we consider
$\re\,(\vW\,A\,B) : (\gamma : \re\,\Gamma) \to \whset$. Externally, the elements
of $\re\,(\vW\,A\,B)\,\gamma$ are CFTT terms with a $\vW$-type in purely
object-level contexts. We call such terms \emph{restricted}. Now, we internalize
the restriction of the $\vsup$ constructor:
\[ \re_{\vsup} : (\alpha : \re\,A\,\gamma) \to \re\,(B \Rightarrow (W\,A\,B)[\p])\,(\gamma,\,\alpha) \to \re\,(W\,A\,B)\,\gamma \]
Externally, $\re_{\vsup}$ applies $\vsup$ to restricted terms. Note that in
$\re\,(B \Rightarrow (W\,A\,B)[\p])\,(\gamma,\,\alpha)$, the type $B \Rightarrow (W\,A\,B)[\p]$ is in
the context $(\Gamma\,\ext\,A)$, which makes it possible to pass $(\gamma,\,\alpha)$ as
argument; this ``instantiates'' $B$ to $\alpha$. With this, we can also
show that $\re$ preserves $\vsup$:
\[ \re\,(\vsup\,t\,u)\,\gamma = \re_\vsup\,(\re\,t\,\gamma)\,(\re\,u\,\gamma) \]
Note that the type of $\re\,u\,\gamma$ is $\re\,(B[\id,\,t] \Rightarrow
\vW\,A\,B)\,\gamma$, which is equal to $\re\,((B \Rightarrow
(\vW\,A\,B)[\p])[\id,\,t])\,\gamma$, which is in turn equal to $\re\,(B
\Rightarrow (\vW\,A\,B)[\p])\,(\gamma,\,\re\,t\,\gamma)$ by $\re$'s preservation
of substitutions, which makes the above equation well-typed.

>>>>

%% \documentclass[acmsmall,screen,review,anonymous]{acmart}
%% \documentclass[nonacm,acmsmall]{acmart}
\documentclass[acmsmall]{acmart}
%% \raggedbottom

@@

\setcopyright{acmlicensed}
\acmDOI{10.1145/3674648}
\acmYear{2024}
\copyrightyear{2024}
\acmSubmissionID{icfp24main-p75-p}
\acmJournal{PACMPL}
\acmVolume{8}
\acmNumber{ICFP}
\acmArticle{259}
\acmMonth{8}
\received{2024-02-28}
\received[accepted]{2024-06-18}

@@

\title{Closure-Free Functional Programming in a Two-Level Type Theory}
\titlenote{This work was supported
by grant 2019.0116 of the Knut and Alice Wallenberg Foundation.}

\author{András Kovács}
\orcid{0000-0002-6375-9781}
\affiliation{%
  \institution{University of Gothenburg}
  \city{Gothenburg}
  \country{Sweden}
}
\email{andrask@chalmers.se}

@@

\section*{Acknowledgements}
I thank the anonymous reviewers, Sebastian Graf and Jacques Carette, for their
comments and suggestions on draft versions of this paper.

@@

%% --------------------------------------------------------------------------------
%% --------------------------------------------------------------------------------

@@

\subsection{W-types}
So far we have stayed close to \cite{staged2ltt}, with only cosmetic
changes. Supporting W-types, as a way to get inductive families \cite{whynotw},
is the biggest extension here. We extend Section 5 of \cite{staged2ltt}. The
formal specification of syntactic $\vW{-}\msf{elim}$ is rather gnarly, because
of the usage of explicit substitutions, but we write it out here anyways.

\begin{notation} We use $\p^N$ for $N$-fold
weakening and we also use numerals as De Bruijn indices, to abbreviate
$\q[\p^N]$. We write $\msf{app}$ and $\msf{lam}$ for $\Pi$ application and
abstraction. We write $\blank{\Rightarrow}\blank$ for non-dependent function
types. Finally, for the sake of brevity we omit the component maps of the isomorphism
$\re_\ext : \re\,(\Gamma \ext A) \simeq ((\gamma : \re\,\Gamma) \times \re\,A\,\gamma)$.
\end{notation}

In the syntax of CFTT, $\vW$-types are as follows.
\begin{alignat*}{3}
  &\vW   &&: (A : \MTy_i\,\Gamma) \to \MTy_j\,(\Gamma\,\ext\,A) \to \MTy_{\msf{max}(i,j)}\,\Gamma\\
  &\vsup &&: (a : \MTm\,\Gamma\,A) \to \MTm\,\Gamma\,(B[\vid,\,a] \Rightarrow \vW\,A\,B) \to \MTm\,\Gamma\,(\vW\,A\,B)\\
  &\vW{-}\msf{elim}  &&: (P : \MTy\,(\Gamma\,\ext\,\vW\,A\,B))\\
  &                  && \hspace{0.6em}\to \MTm\,(\Gamma\,\ext\,A\,\ext\,B \Rightarrow (\vW\,A\,B)[p]\,\ext\,\Pi\,(B[p])\,(P[\p^3,\,\msf{app}\,1\,0]))\,(P[\p^3,\,\vsup\,2\,1))\\
  &                  && \hspace{0.6em}\to (w : \MTm\,\Gamma\,(\vW\,A\,B)) \to \MTm\,\Gamma\,(P[\id,\,w])\\
  & \vW{\beta}       &&:
    \vW{-}\msf{elim}\,P\,s\,(\vsup\,a\,f) = s[\id,\,a,\,f,\,\msf{lam}\,(\vW{-}\msf{elim}\,(P[\p^2,\,\q])\,(s[\p])\,(\msf{app}\,f\,0)]
\end{alignat*}
We shall omit the $i$ and $j$ universe levels in the following. In $\whset$,
i.e.\ internally to presheaves over $\mbbo$, we also have the standard W-type.
\begin{alignat*}{3}
  & \wh{\vW}   &&: (A : \whset) \to (A \to \whset) \to \whset\\
  & \wh{\vsup} &&: (a : A) \to (B\,a \to \wh{\vW}\,A\,B) \to \wh{\vW}\,A\,B\\
  & \wh{\vW{-}\msf{elim}} &&: (P : \wh{\vW}\,A\,B \to \whset)\\
  & && \hspace{0.6em}\to ((a : A)(f : B\,a \to \wh{\vW}\,A\,B) \to ((b : B\,a) \to P\,(f\,b)) \to P\,(\wh{\vsup}\,a\,f))\\
  & && \hspace{0.6em}\to (w : \wh{\vW}\,A\,B) \to P\,w\\
  & \wh{\vW}{\beta} &&: \wh{\vW{-}\msf{elim}}\,P\,s\,(\wh{\vsup}\,a\,f) = s\,a\,f\,(\lambda\,b.\,\wh{\vW{-}\msf{elim}}\,P\,s\,(f\,b))
\end{alignat*}
In the presheaf model we interpret $\vW$ using $\whW$.
\begin{alignat*}{3}
  & \ev\,(\vW\,A\,B)\,\gamma   &&:= \wh{\vW}\,(\ev\,A\,\gamma)\,(\lambda\,\alpha.\,\ev\,B\,(\gamma,\,\alpha))\\
  & \ev\,(\vsup\,a\,f)\,\gamma &&:= \wh{\vsup}\,(\ev\,a\,\gamma)\,(\lambda\,\beta.\,\ev\,f\,(\gamma, \beta))\\
  & \ev\,(\vW{-}\msf{elim}\,P\,s\,w)\,\gamma &&:= \wh{\vW{-}\msf{elim}}\,(\lambda\,w.\,\ev\,P\,(\gamma,\,w))\,(\lambda\,a\,f\,g.\,\ev\,s\,(\gamma,\,a,\,f,\,g))\,(\ev\,w\,\gamma)
\end{alignat*}

%% W-elim̂ (λ w. E P (γ, w)) (λ a f g. E s (γ, a, f, g)) (E w γ)
Now, assuming $A : \MTy\,\Gamma$ and $B : \MTy\,(\Gamma\,\ext\,A)$, we consider
$\re\,(\vW\,A\,B) : (\gamma : \re\,\Gamma) \to \whset$. Externally, the elements
of $\re\,(\vW\,A\,B)\,\gamma$ are CFTT terms with a $\vW$-type in purely
object-level contexts. We call such terms \emph{restricted}. Now, we internalize
the restriction of the $\vsup$ constructor:
\[ \re_{\vsup} : (\alpha : \re\,A\,\gamma) \to \re\,(B \Rightarrow (W\,A\,B)[\p])\,(\gamma,\,\alpha) \to \re\,(W\,A\,B)\,\gamma \]
Externally, $\re_{\vsup}$ applies $\vsup$ to restricted terms. Note that in
$\re\,(B \Rightarrow (W\,A\,B)[\p])\,(\gamma,\,\alpha)$, the type $B \Rightarrow (W\,A\,B)[\p]$ is in
the context $(\Gamma\,\ext\,A)$, which makes it possible to pass $(\gamma,\,\alpha)$ as
argument; this ``instantiates'' $B$ to $\alpha$. With this, we can also
show that $\re$ preserves $\vsup$:
\[ \re\,(\vsup\,t\,u)\,\gamma = \re_\vsup\,(\re\,t\,\gamma)\,(\re\,u\,\gamma) \]
Note that the type of $\re\,u\,\gamma$ is $\re\,(B[\id,\,t] \Rightarrow
\vW\,A\,B)\,\gamma$, which is equal to
\[\re\,((B \Rightarrow(\vW\,A\,B)[\p])[\id,\,t])\,\gamma\]
which is in turn equal to $\re\,(B \Rightarrow
(\vW\,A\,B)[\p])\,(\gamma,\,\re\,t\,\gamma)$ by $\re$'s preservation of
substitutions, which makes the above equation well-typed.
